---
title: "Model Fitting Exercise"
author: Prasanga Paudel
date: February 28, 2025
format:
  html:
    toc: false
    number-sections: true
    highlight-style: github
---


```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(dplyr)
library(ggplot2)
library(here)
library(tidyr)
library(knitr)
library(patchwork)
library(GGally)
library(caret)
library(pROC)
```



# Importing Dataset

We will first import our dataset from the directory.

```{r}
# Constructing the file path using here()
file_path <- here("fitting-exercise","data", "Mavoglurant_A2121_nmpk.csv")

# Importing the Excel file from the file path
Mavoglurant_data <- read.csv(file_path)

# Viewing the uppermost data of the imported dataset
head(Mavoglurant_data)
```




# Looking into the dataset and preparing for EDA 


In this section, we will have an initial look into the dataset and prepare a new dataset for EDA.


## Plotting DV on Y-axis and TIME on X-axis

Here, we will write a code to make a plot that shows a line for each individual, with DV on the y-axis and time on the x-axis, stratified by dose ( using a different color for each dose). 

```{r}
# Creating a line plot with different colors for each dose
ggplot(Mavoglurant_data, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  geom_line() +
  labs(title = "DV over Time by ID",
       x = "Time",
       y = "DV",
       color = "Dose") +
  theme_minimal()
```

_This figures show the three doses in three colors. The DOSE 37.5 is not visible in this graph, probably because it is less frequent._


## Preparing the data

As there are some individuals that seem to have received the drug more than once, we will only keep observations with OCC = 1

```{r}
#writing a code that keeps only observations with OCC = 1
Mavoglurant_data_OCC1 <- subset(Mavoglurant_data, OCC == 1)

#writing code to exclude the observations with TIME = 0
Mavoglurant_data_filtered <- Mavoglurant_data_OCC1 %>%
  filter(TIME != 0)
```


## Computing the total amount of drug for each individual by adding all the DV values

Here, we will compute the total amount of drug for each individual by adding all the DV values.

```{r}
summarized_data <- Mavoglurant_data_filtered %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE))


dim(summarized_data) # checking dimension
```


Now, we will create a dataframe that contains only the observations where TIME == 0, and then we will check the dimensions to see if it is correctly created. 
```{r}
# creating a data frame that contains only the observations where TIME == 0
time_zero_data <- Mavoglurant_data %>%
  filter(TIME == 0)

dim(time_zero_data) # checking dimension
```

Now, we will create a new dataframe using the join function and then we will check the dimensions to see if it is correctly created. 
```{r}
# using the join function to combine those two data frames
final_data <- time_zero_data %>%
  left_join(summarized_data, by = "ID")


dim(final_data) # checking dimension
```

## Creating the final dataset

Writing a code that converts RACE and SEX to factor variables and keeps only these variables: Y,DOSE,AGE,SEX,RACE,WT,HT
```{r}
final_data_cleaned <- final_data %>% mutate (
  RACE = factor(RACE), SEX = factor(SEX)) %>% select (Y, DOSE, AGE, SEX, RACE, WT, HT
)

dim(final_data_cleaned)
```


# Exploratory Data Analysis

Now, we will go through the Exploratory Data Analysis process. We will create tables, figures, and plots to see if there is any relation between the variables. 

##Making some useful summary tables



Here, we create a summary table of all the variables we have in our final dataset.

```{r}

# Calculating summary statistics for variables
summary_stats <- final_data_cleaned %>%
  summarise(across(c(Y, DOSE, AGE, WT, HT), 
                 list(Mean = ~ mean(., na.rm = TRUE),
    Min = ~ min(., na.rm = TRUE),
    Max = ~ max(., na.rm = TRUE),
    SD = ~ sd(., na.rm = TRUE))))

# Reshaping the data to better present it
summary_stats <- summary_stats %>%
  pivot_longer(cols = everything(), 
               names_to = c("Variable", ".value"), 
               names_sep = "_")

# Printing the summary statistics as a  table format
kable(summary_stats, caption = "Summary Statistics", align = "c")
```

_This table represennts the summary statistics of the 5 variables._



## Plotting the relationship with outcome of interest (Y)


Here, we will show some scatterplots or boxplots, whichever is suitable, between the main outcome of interest (total drug, Y) and other predictors. We will use box-plot if the variable is categorical.


```{r}

# Scatterplot: Y vs DOSE
p1 <- ggplot(final_data_cleaned, aes(x = DOSE, y = Y)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Y vs DOSE", x = "DOSE", y = "Y")

# Scatterplot: Y vs AGE
p2 <- ggplot(final_data_cleaned, aes(x = AGE, y = Y)) +
  geom_point(alpha = 0.6, color = "green") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Y vs AGE", x = "AGE", y = "Y")

# Scatterplot: Y vs WT
p3 <- ggplot(final_data_cleaned, aes(x = WT, y = Y)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Y vs WT", x = "WT", y = "Y")

# Scatterplot: Y vs HT
p4 <- ggplot(final_data_cleaned, aes(x = HT, y = Y)) +
  geom_point(alpha = 0.6, color = "orange") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Y vs HT", x = "HT", y = "Y")

# Boxplot: Y vs SEX
p5 <- ggplot(final_data_cleaned, aes(x = SEX, y = Y, fill = SEX)) +
  geom_boxplot() +
  labs(title = "Y vs SEX", x = "SEX", y = "Y") +
  theme(legend.position = "none")

# Boxplot: Y vs RACE
p6 <- ggplot(final_data_cleaned, aes(x = RACE, y = Y, fill = RACE)) +
  geom_boxplot() +
  labs(title = "Y vs RACE", x = "RACE", y = "Y") +
  theme(legend.position = "none")

# Combine all plots into a common canvas
combined_plots <- (p1 + p2) / (p3 + p4) / (p5 + p6)

# Display the combined plots
combined_plots

```
We can observe from the figures above that: 

1. Dose and Y (Total drug) dont have a strong correlation.
2. Age and Y dont have a strong correlation.
3. Weight and Y have a negative correlation.
4. Height and Y have a negative correlation.
5. Sex 1 on average has higher total drug.

## Plotting the distribution of variables

Now, we will plot the distributions of our variables to make sure they all make sense. We will set the bin according to the values and distribution patterns within the variable 

```{r}

# Custom theme for consistent styling
custom_theme <- theme(
  plot.title = element_text(size = 14, face = "bold"),
  axis.title = element_text(size = 12),
  axis.text = element_text(size = 10),
  plot.margin = margin(1, 1, 1, 1, "cm")
)

# Histogram for Y
p1 <- ggplot(final_data_cleaned, aes(x = Y)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "black") +
  labs(title = "Distribution of Y", x = "Y", y = "Count") +
  custom_theme

# Histogram for DOSE
p2 <- ggplot(final_data_cleaned, aes(x = DOSE)) +
  geom_histogram(binwidth = 10, fill = "green", color = "black") +
  labs(title = "Distribution of DOSE", x = "DOSE", y = "Count") +
  custom_theme

# Histogram for AGE
p3 <- ggplot(final_data_cleaned, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "purple", color = "black") +
  labs(title = "Distribution of AGE", x = "AGE", y = "Count") +
  custom_theme

# Histogram for WT
p4 <- ggplot(final_data_cleaned, aes(x = WT)) +
  geom_histogram(binwidth = 5, fill = "orange", color = "black") +
  labs(title = "Distribution of WT", x = "WT", y = "Count") +
  custom_theme

# Histogram for HT
p5 <- ggplot(final_data_cleaned, aes(x = HT)) +
  geom_histogram(binwidth = 0.01, fill = "red", color = "black") +
  labs(title = "Distribution of HT", x = "HT", y = "Count") +
  custom_theme

# Bar plot for SEX
p6 <- ggplot(final_data_cleaned, aes(x = SEX)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Distribution of SEX", x = "SEX", y = "Count") +
  custom_theme

# Bar plot for RACE
p7 <- ggplot(final_data_cleaned, aes(x = RACE)) +
  geom_bar(fill = "pink", color = "black") +
  labs(title = "Distribution of RACE", x = "RACE", y = "Count") +
  custom_theme

# Combining all plots into a common canvas
combined_plots <- (p1 + p2) / (p3 + p4) / (p5 + p6 + p7)

# Display the combined plots
combined_plots


```

The distribution of the variables make sense in my opinion. Age, Weight, Height are all fairly normally distributed. Total drug has a somewhat uniform distribution. Other categorical variables have an unbalanced distribution among them but it is practically possible if the research is focused on specific group of people.

## Pair-wise plots

Here, we will plot different variables against each other to see any possible relation or pattern between the variables. 

_Note: Please ignore the plots of a variable against itself as it does not make any sense._

```{r}

# Selecting the variables of interest
selected_vars <- final_data_cleaned %>%
  select(Y, DOSE, AGE, WT, HT)  # to nclude only numeric variables for pair plots

# Creating a pair plot
pair_plot <- ggpairs(selected_vars, 
                     title = " Plotting pairs of Numeric Variables",
                     lower = list(continuous = wrap("smooth", alpha = 0.3, color = "blue")),  # Add smooth lines
                     diag = list(continuous = wrap("densityDiag", fill = "orange")),  # Density plots on diagonal
                     upper = list(continuous = wrap("cor", size = 4)))  # Add correlation coefficients

# Disply the pair plot
print(pair_plot)


```
It is visible that HT and WT are strongly correlated. Y shows correlation with WT. Age and WT also show some moderate correlation.

## Correlation among variables

Here, we will create a cross-table matrix that shows the correlation among our variable of interests.

```{r}
# Creating a correlation plot
cor_plot <- ggcorr(selected_vars, 
                   method = c("pairwise", "pearson"),  # Use Pearson correlation
                   label = TRUE,  # to display correlation values
                   label_size = 4, 
                   color = "grey50", 
                   hjust = 0.75, 
                   size = 3, 
                   layout.exp = 1)

# Display the correlation plot
print(cor_plot)

```

It is visible that HT and WT show the maximum correlation with a correlation coefficient of 0.6. Age and HT also show sloghtly weaker correlation. Dose has no correlation with WT and HT.

# Fitting a model for Y with only DOSE, and ALL predictors 


We will now fit a linear model to the continuous outcome (Y) using the main predictor of interest to be DOSE in one case and using all predictors in another case.

## Fitting a linear model using only DOSE as the predictor


```{r}
# Fitting a linear model using only DOSE as the predictor
model_dose <- lm(Y ~ DOSE, data = final_data_cleaned)

summary(model_dose)
```
We can observe that dose has an insignificant effect on Y.



## Fitting a linear model using all predictor


```{r}
# Fit a linear model using all predictors
model_all <- lm(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = final_data_cleaned)

summary(model_all)
```
We can observe that SEX2 and Wt has a significant effect on Y. But majority o the predictors are insignificant in the model. This has lead to an insufficient information to predict the dependent vriable and only around 10% variation in the dependent variable is explained by the model with all predictors. With only DOSE the performance is even poorer.



## computing RMSE and R-squared for both models

Here, we will compute the RMSE and R-squared for the both the models discussed above. 


```{r}


# creating function to compute RMSE and R-squared
compute_metrics <- function(model, data) {
  # Predictions
  predictions <- predict(model, newdata = data)
  
  # Actual values
  actual <- data$Y
  
  # Computing RMSE
  rmse <- sqrt(mean((actual - predictions)^2))
  
  # Compute R-squared
  r_squared <- summary(model)$r.squared
  

  return(list(RMSE = rmse, R_squared = r_squared))
}

# Computing metrics for the DOSE-only model
metrics_dose <- compute_metrics(model_dose, final_data_cleaned)

# Computing metrics for the all-predictors model
metrics_all <- compute_metrics(model_all, final_data_cleaned)

# Creatin a data frame with the results
results_table <- data.frame(
  Model = c("DOSE Only", "All Predictors"),
  RMSE = c(metrics_dose$RMSE, metrics_all$RMSE),
  R_squared = c(metrics_dose$R_squared, metrics_all$R_squared)
)

kable(results_table, caption = "Model Performance Metrics", align = "c")

```

_This tables shows the model performance with Y as dependent variable._



# Fitting a Logistic model for SEX with only DOSE, and ALL predictors 


We will now fit a linear model to the binary outcome (SEX) using the main predictor of interest to be DOSE in one case and using all predictors in another.


## Fiting a logistic model using only DOSE as the predictor


```{r}
# Ensuring SEX is a binary factor
final_data_cleaned <- final_data_cleaned %>%
  mutate(SEX = factor(SEX))

# Fiting a logistic model using only DOSE as the predictor
logistic_dose <- glm(SEX ~ DOSE, data = final_data_cleaned, family = binomial())
summary(logistic_dose)
```
We can observe that DOSE does not significantly affect SEX.


## Fitting a logistic model using all predictors

```{r}
# Fit a logistic model using all predictors
logistic_all <- glm(SEX ~ DOSE + AGE + RACE + WT + HT + Y, data = final_data_cleaned, family = binomial())


summary(logistic_all)
```
We can observe that HT and Y significantly affect Y. We will need to check the Accuracy and AUC values to tell about the performance of the model.


## Computing accuracy and ROC-AUC for both the models

Here, we will compute  accuracy and ROC-AUC for both the models discussed above with SEX as dependent variable.

```{r}

# Function to compute accuracy and ROC-AUC
compute_metrics_logistic <- function(model, data) {
  # Predict probabilities
  predicted_probs <- predict(model, newdata = data, type = "response")
  
  # Convert probabilities to binary predictions (0 or 1)
  predicted_classes <- ifelse(predicted_probs > 0.5, 1, 0)
  
  # Actual classes
  actual_classes <- as.numeric(data$SEX) - 1  # Convert factor to numeric (0 or 1)
  
  # Compute accuracy
  accuracy <- mean(predicted_classes == actual_classes)
  
  # Compute ROC-AUC
  roc_auc <- roc(actual_classes, predicted_probs)$auc
  
  # Return metrics
  return(list(Accuracy = accuracy, ROC_AUC = roc_auc))
}

# Compute metrics for the DOSE-only model
metrics_dose <- compute_metrics_logistic(logistic_dose, final_data_cleaned)

# Compute metrics for the all-predictors model
metrics_all <- compute_metrics_logistic(logistic_all, final_data_cleaned)

results_table_logistic <- data.frame(
  Model = c("DOSE Only", "All Predictors"),
  Accuracy = c(metrics_dose$Accuracy, metrics_all$Accuracy),
  ROC_AUC = c(metrics_dose$ROC_AUC, metrics_all$ROC_AUC)
)

# Print the table using kable
kable(results_table_logistic, caption = "Logistic Model Performance Metrics", align = "c")
```
The tablbe above shoes the  accuracy and ROC-AUC for both the models discussed above with SEX as dependent variable. The AUC value for DOSE only is similar to a random model. The AUC value for All predictors is close to 1 implying a nearly perfect prediction performance. I personally find this very surprising that we can predict SEX so perfectly.

